---
title: "Code_Documentation"
author: "Hasanhejleh"
date: "2025-03-12"
output: 
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
    fig_caption: true
    highlight: tango
---

# Divvy Bike Share Data Analysis

## **Project Overview**

This analysis explores the Divvy bike share dataset to identify trends in bike usage, membership types, and station popularity. The dataset includes trip data from January 2024 to January 2025, with details such as trip duration, distance, start/end stations, and user type (casual vs. annual members).

## **Setting Up The Environment**

### 1. Install&Load Required Packages.
Before starting the analysis, ensure the necessary R packages are installed. These packages include `tidyverse`, `skimr`, `janitor`, `dplyr`, and `geosphere`.

```{r install-packages}
#install.packages("tidyverse")
#install.packages("skimr")
#install.packages("janitor")
#install.packages("dplyr")
#install.packages("geosphere")
#install.packages("tinytex")
#tinytex::install_tinytex()
```

```{r load-packages}
library(tidyverse)
library(skimr)
library(janitor)
library(dplyr)
library(geosphere)
library(tinytex)
```

## **Loading and Combining Data**

### 1. Load Monthly Trip Data

The Divvy trip data is stored in separate CSV files for each month. Load each file into a dataframe.

```{r load-data}
divvy2024_01_df <- read_csv("202401-divvy-tripdata.csv")
divvy2024_02_df <- read_csv("202402-divvy-tripdata.csv")
divvy2024_03_df <- read_csv("202403-divvy-tripdata.csv")
divvy2024_04_df <- read_csv("202404-divvy-tripdata.csv")
divvy2024_05_df <- read_csv("202405-divvy-tripdata.csv")
divvy2024_06_df <- read_csv("202406-divvy-tripdata.csv")
divvy2024_07_df <- read_csv("202407-divvy-tripdata.csv")
divvy2024_08_df <- read_csv("202408-divvy-tripdata.csv")
divvy2024_09_df <- read_csv("202409-divvy-tripdata.csv")
divvy2024_10_df <- read_csv("202410-divvy-tripdata.csv")
divvy2024_11_df <- read_csv("202411-divvy-tripdata.csv")
divvy2024_12_df <- read_csv("202412-divvy-tripdata.csv")
divvy2025_01_df <- read_csv("202501-divvy-tripdata.csv")
```

### 2. Check Data Integrity

Verify the structure and integrity of each dataframe by inspecting column names, the first few rows, and the data types.

-Repeat this step for all 13 dataframes.

```{r check-data}
colnames(divvy2024_01_df)
head(divvy2024_01_df)
str(divvy2024_01_df)
```

### 3. Combine Data into a Single Dataframe

Aggregate all monthly data into a single dataframe for easier analysis.

-Note: change the path to the file you have the data inside.

```{r combine-data}
csv_files <- list.files(path = "D:/Moved Data/Google Data Analytics Program/Case Study Stuff/R Studio code",
                        pattern = "*.csv",
                        full.names = TRUE)

divvy_combined_df <- csv_files %>% 
  lapply(function(file) {
    
    df <- read_csv(file)
    df$yearmonth <- gsub(pattern = "[^0-9]", replacement = "",
                         basename(file))
    
    return(df)
    
  }) %>% 
  bind_rows()
```

## **Data Cleaning and Pre processing**

### **Data Cleaning Steps**

1. **Reformat Columns**:
   - Convert `start_lat` and `end_lat` to numeric.
   - Convert `start_lng` and `end_lng` to numeric.

2. **Handle Missing Values**:
   - Remove rows with missing `end_lat` and `end_lng`.
   - Replace missing station names and IDs with "Unknown".

3. **Create New Columns**:
   - Calculate trip duration.
   - Calculate trip distance using the Haversine formula.

4. **Filter Invalid Trips**:
   - Remove trips with duration < 1.3 minutes.
   - Remove trips with distance < 200 meters or > 768 kilometers.

5. **Remove Duplicates**:
   - Round timestamps to the nearest minute.
   - Remove duplicate rows based on `ride_id`, `started_at`, and `ended_at`.


### 1. **Reformat Columns**

Ensure latitude and longitude columns are in the correct numeric format.

```{r reformat-columns}
divvy_combined_df$start_lat <- as.double(divvy_combined_df$start_lat)
divvy_combined_df$end_lat <- as.double(divvy_combined_df$end_lat)
divvy_combined_df$start_lng <- as.double(divvy_combined_df$start_lng)
divvy_combined_df$end_lng <- as.double(divvy_combined_df$end_lng)
```

### 2. **Handle Missing Values**

Check for and address missing values in the dataset.

```{r NA-missing-unknown-values}
colSums(is.na(divvy_combined_df))

# Remove NA values for end_lat and end_lng
divvy_combined_df <- divvy_combined_df %>% 
  filter(!is.na(end_lat))

# Replace NA with "Unknown" for station names and IDs
divvy_combined_df$start_station_name[is.na(divvy_combined_df$start_station_name)] <- "Unknown"
divvy_combined_df$start_station_id[is.na(divvy_combined_df$start_station_id)] <- "Unknown"
divvy_combined_df$end_station_name[is.na(divvy_combined_df$end_station_name)] <- "Unknown"
divvy_combined_df$end_station_id[is.na(divvy_combined_df$end_station_id)] <- "Unknown"

# Recheck for NA values
colSums(is.na(divvy_combined_df))

```

### 3. **Create New Columns**

- Create Trip Duration Column

Calculate the duration of each trip in minutes.


```{r trip-duration-column}
divvy_combined_df$started_at <- as.POSIXct(divvy_combined_df$started_at, format = "%Y-%m-%d %H:%M:%S", tz = "America/Chicago")
divvy_combined_df$ended_at <- as.POSIXct(divvy_combined_df$ended_at, format = "%Y-%m-%d %H:%M:%S", tz = "America/Chicago")

duration <- difftime(divvy_combined_df$ended_at, divvy_combined_df$started_at, units = "mins")
divvy_combined_df <- divvy_combined_df %>% mutate(duration)
```

- Create Trip Distance Column

Calculate the distance of each trip using the Haversine formula.

```{r trip-distance-column}
divvy_combined_df <- divvy_combined_df %>% 
  mutate(distance = distHaversine(cbind(start_lng, start_lat), cbind(end_lng, end_lat)))
```

- Create Weekday Column

```{r start-weekday-column}

# Create new start_weekday column

divvy_combined_df <- divvy_combined_df %>% 
  mutate(start_weekday = weekdays(started_at))
```


### 4. **Filter Invalid Trips**

Remove trips with unrealistic durations or distances, and calculate the percentage of those trips.

```{r filter-invalid-trips}

percent_short_trips <- mean(divvy_combined_df$distance < 200 | divvy_combined_df$duration < 1.30 | divvy_combined_df$distance > 768000) * 100
print(paste("Percentage of trips < 200 meters OR < 1.30 minutes OR > 768000 meters:", round(percent_short_trips, 2), "%"))

invalid_trips <- divvy_combined_df %>% 
  filter(duration < 1.30 & (distance < 200 | distance > 768000))

divvy_combined_df <- divvy_combined_df %>% 
  filter(duration >= 1.3 & distance >= 200 & distance <= 768000)

# Recheck

percent_short_trips <- mean(divvy_combined_df$distance < 200 | divvy_combined_df$duration < 1.30 | divvy_combined_df$distance > 768000) * 100
print(paste("Percentage of trips < 200 meters OR < 1.30 minutes OR > 768000 meters:", round(percent_short_trips, 2), "%"))
```

### 5. **Remove Duplicates**

Identify and remove duplicate entries based on ride_id, started_at, and ended_at.

```{r remove-duplicates}
duplicates <- divvy_combined_df %>%
  group_by(ride_id) %>%
  filter(n() > 1) %>%
  arrange(ride_id)

divvy_combined_df$started_at <- round_date(divvy_combined_df$started_at, unit = "minute")
divvy_combined_df$ended_at <- round_date(divvy_combined_df$ended_at, unit = "minute")

divvy_combined_df <- divvy_combined_df %>% distinct(ride_id, started_at, ended_at, .keep_all = TRUE)
```

## Data Analysis and Visualization

### 1. Distribution of Trip Durations

Analyze the distribution of trip durations and visualize it.

```{r trip-duration-distribution-viz-1}
# Reformat duration column from timediff object to numeric
divvy_combined_df$duration <- as.numeric(divvy_combined_df$duration)

# Define bins bounds
lower_bound <- min(divvy_combined_df$duration)
upper_bound <- max(divvy_combined_df$duration)

# Define duration bins (adjust based on data distribution)
bins <- c(lower_bound, 5, 10, 30, 60, 120, 300, 600, 900, 1200, upper_bound)


# Compute percentage in each bin
duration_distribution <- divvy_combined_df %>%
  mutate(duration_bin = cut(duration, breaks = bins, include.lowest = TRUE)) %>%
  group_by(duration_bin) %>%
  summarise(count = n()) %>%
  mutate(percentage = (count / sum(count)) * 100) %>%
  arrange(desc(percentage))  # Sort by highest concentration

# View the distribution
print(duration_distribution)

# Visualize the distribution

ggplot(data = divvy_combined_df, aes(x = duration)) +
  geom_histogram(bins = 40, color = "black", fill = "steelblue", alpha = 0.7) +
  scale_x_log10(breaks = c(1.3, 5, 10, 30, 60 ,120))+
  labs(
    title = "Distribution Of Trips By Duration (minutes)",
    x = "Duration (minutes)",
    y = "Frequency"
    
  ) +
  theme_minimal()


```

### 2. Membership Type & Number of Trips

Visualize the relationship between membership type and the number of trips.

```{r membership-type-trips-count-viz-2}

# Group trips by membership type & bike type

avg_duration <- divvy_combined_df %>% 
  group_by(member_casual, rideable_type) %>% 
  summarise(average_duration = mean(duration), trip_count = n(), .groups = "drop") %>%
  arrange(member_casual, rideable_type)

# Create a stacked bar chart visualization for the segments of the data

ggplot(avg_duration, aes(x = member_casual, y = trip_count, fill = rideable_type)) +
  geom_bar(stat = "identity") + 
  geom_text(aes(label = scales::comma(trip_count)), 
            position = position_stack(vjust = 0.5),  
            color = "black", size = 5, fontface = "bold") +
  labs(
    title = "Number of Trips For Each Membership Type",
    x = "User Type",
    y = "Number of Trips",
    caption = "1,924,305 casual members & 3,620,854 annual members"
  ) +
  theme_minimal() +
  theme(plot.caption = element_text(hjust = 1))

```

### 3. Average Duration for Each Bike Type by User Type

Analyze the average trip duration for each membership type.

```{r average-duration-boxplot-viz-3}

# Precompute summary stats (min, max, median) for each member type

summary_data <- avg_duration %>%
  group_by(member_casual) %>%
  summarise(
    min_duration = min(average_duration),
    min_rideable = rideable_type[which.min(average_duration)],
    max_duration = max(average_duration),
    max_rideable = rideable_type[which.max(average_duration)],
    median_duration = median(average_duration),
    median_rideable = rideable_type[which.min(abs(average_duration - median(average_duration)))], # Closest to median
    .groups = "drop"
  )

# Create the boxplot
ggplot(avg_duration, aes(x = member_casual, y = average_duration, fill = member_casual)) +
  geom_boxplot() +
  
  # Add min value text
  geom_text(data = summary_data, aes(x = member_casual, y = min_duration,
                                     label = paste0(round(min_duration, 2), " / ", min_rideable)),
            vjust = 1.5, color = "black", size = 3) +
  
  # Add max value text
  geom_text(data = summary_data, aes(x = member_casual, y = max_duration,
                                     label = paste0(round(max_duration, 2), " / ", max_rideable)),
            vjust = -0.5, color = "black", size = 3) +
  
  # Add median value text
  geom_text(data = summary_data, aes(x = member_casual, y = median_duration,
                                     label = paste0(round(median_duration, 2), " / ", median_rideable)),
            vjust = -0.2, color = "black", size = 3) +
  
  labs(title = "Average Duration for Each Bike Type by User Type",
       x = "User Type",
       y = "Average Duration (minutes)") +
  theme_minimal()

```

### 4. Casual Vs Annual Members Trips Distribution On Weekdays

Analyze which days of the week riders prefer to take trips.

```{r trips-distribution-weekdays-viz-4}

# Day of the week relationship & number of rides

divvy_combined_df %>% 
  group_by(start_weekday) %>% 
  summarise(num_of_rides = n()) %>% 
  arrange(-num_of_rides)


# Visualize the relationship using a stacked bar chart with bike type segments

ggplot(data = divvy_combined_df, mapping = aes(x = start_weekday, fill = rideable_type)) +
  geom_bar() + 
  facet_wrap(~member_casual) + 
  labs(
    x = "Week Days",
    y = "Number Of Trips",
    title = "Casual Vs Annual Members Trips Distribution On Weekdays"
  ) +
  theme_minimal() +
  theme(
    # Add margin between facets
    panel.spacing = unit(1.5, "lines"),  # Increase spacing between facets
    
    # Adjust x-axis text angle, spacing, and alignment
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10, margin = margin(t = 10)),  # Rotate x-axis text and add margin
    
    # Add padding to the plot
    plot.margin = margin(1, 1, 1, 1, "cm"),  # Add margin around the plot
    
    # Adjust facet strip text
    strip.text = element_text(size = 12, face = "bold")  # Customize facet labels
  )

```

### 5. Number of Trips by Hour

Analyze when riders tend to use bikes throughout the day.


```{r trips-by-hour-viz-5}

# Add a start_hour column by extracting the hour from started_at column

divvy_combined_df <- divvy_combined_df %>% 
  mutate(start_hour = format(as.POSIXct(started_at), "%H"))

# Create a data frame to store the number of trips and their relative start hours

trips_by_hour <- divvy_combined_df %>%
  group_by(start_hour) %>%
  summarise(num_of_trips = n())

# Add a new formatted column for the start_hour values in the visualization

trips_by_hour <- trips_by_hour %>%
  mutate(start_hour_numeric = as.numeric(sub(":00:00", "", start_hour)))

ggplot(trips_by_hour, aes(x = start_hour_numeric, y = num_of_trips)) +
  geom_point() +
  geom_smooth(method = "loess", se = TRUE, color = "blue") +
  scale_x_continuous(breaks = seq(0, 23, by = 1), labels = paste0(seq(0, 23))) +
  labs(title = "Number of Trips by Hour",
       x = "Hour of the Day (24-hour format)",
       y = "Number of Trips") +
  theme_minimal()
```


### 6. Busiest Stations by Membership Type

Identify and visualize the top 10 busiest stations for casual and annual members.

```{r start_station_counts-creation}
# Data Frame that includes: station names, membership type ,number of trips

start_station_counts <- divvy_combined_df %>%
  group_by(start_station_name, member_casual) %>%
  summarise(num_of_trips = n(), .groups = 'drop') %>%
  arrange(-num_of_trips)


```


- Top 10 Busiest Casual stations Viz

```{r busiest-casual-stations-viz-5}

# Filter stations by membership type since some stations were only used by 1 membership type

casual_stations <- start_station_counts %>% 
  filter(member_casual == "casual", start_station_name != "Unknown") %>% 
  arrange(-num_of_trips) %>% 
  top_n(10)

# Top 10 Busiest Stations for Casual Members Visualization

options(repr.plot.width = 10, repr.plot.height = 8)

ggplot(data = casual_stations, aes(x = reorder(start_station_name, num_of_trips), y = num_of_trips)) +
  geom_bar(stat = "identity", col = "steelblue", fill = "steelblue") +  # Color bars
  coord_flip() +  # Flip coordinates to make the bars horizontal
  labs(
    x = "Station Name",
    y = "Number of Trips",
    title = "Top 10 Busiest Stations for Casual Members"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 12, family = "sans"),  # Improve x-axis text
    axis.text.y = element_text(size = 12, family = "sans"),  # Improve y-axis text
    axis.title.x = element_text(size = 14, face = "bold", family = "sans"),  # Improve x-axis title
    axis.title.y = element_text(size = 14, face = "bold", family = "sans"),  # Improve y-axis title
    plot.title = element_text(size = 16, face = "bold", family = "sans", hjust = 0.5),  # Improve plot title
    panel.grid.major = element_line(color = "gray80"),  # Add light grid lines for readability
    panel.grid.minor = element_blank(), # Remove minor grid lines
    plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm")  # Adjust margins
  )

```



- Top 10 Busiest Annual stations Viz

```{r busiest-casual-stations-viz-6}

# Filter stations by membership type since some stations were only used by 1 membership type

member_stations <- start_station_counts %>% 
  filter(member_casual == "member", start_station_name != "Unknown") %>% 
  arrange(-num_of_trips) %>% 
  top_n(10)

# Top 10 Busiest Stations for Annual Members Visualization

options(repr.plot.width = 10, repr.plot.height = 8)

ggplot(data = member_stations, aes(x = reorder(start_station_name, num_of_trips), y = num_of_trips)) +
  geom_bar(stat = "identity", col = "#ff7f0e", fill = "#ff7f0e") +  # Color bars
  coord_flip() +  # Flip coordinates to make the bars horizontal
  labs(
    x = "Station Name",
    y = "Number of Trips",
    title = "Top 10 Busiest Stations for Annual Members"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 12, family = "sans"),  # Improve x-axis text
    axis.text.y = element_text(size = 12, family = "sans"),  # Improve y-axis text
    axis.title.x = element_text(size = 14, face = "bold", family = "sans"),  # Improve x-axis title
    axis.title.y = element_text(size = 14, face = "bold", family = "sans"),  # Improve y-axis title
    plot.title = element_text(size = 16, face = "bold", family = "sans", hjust = 0.5),  # Improve plot title
    panel.grid.major = element_line(color = "gray80"),  # Add light grid lines for readability
    panel.grid.minor = element_blank(), # Remove minor grid lines
    plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm")  # Adjust margins
  )

```
